MYC overrides HIF to regulate proliferating primary cell metabolism in hypoxia

### 2022-07-14 manuscript/manuscript.Rmd
start writing up isotope tracing studies

## Hypoxia uncouples glycolytic gene expression from glycolytic metabolic flux

The goal of this study was to characterize hypoxia-induced metabolic changes in proliferating primary LFs and PASMCs. To accomplish this goal, we adopted a metabolic flux analysis (MFA) technique that enabled us to link intracellular metabolic fluxes to cell proliferation rates. MFA fits cell proliferation rate, extracellular flux measurements, and ^13^C intracellular isotope labeling patterns to a model of cell metabolism [@antoniewicz2018]. This analysis reconstructs comprehensive flux maps that depict the flow of carbon from extracellular substrates, throughout intracellular metabolism, and into cell biomass and metabolic by-products [@young2014]. These models assume that cells are at a metabolic pseudo-steady state over the experimental time course [@buescher2015]. Exponential growth phase is thought to reflect metabolic pseudo-steady state, because cells in the culture steadily divide at their maximal condition-specific rate, as long as nutrient supply does not become limiting [@buescher2015; @ahn2011]. Thus, we first set out to define experimental conditions to capture exponential growth phase in normoxic and hypoxic cultures.

@@

## HIF stabilization in normoxia does increase glycolysis

Given that hypoxia exposure did not increase glycolysis in LFs, we next wanted to determine how these cells responded to HIF-1α stabilization in normoxia. To accomplish this, LFs were treated with the PHD inhibitor molidustat (BAY) using a similar time course as our hypoxia experiments. Cells were treated with BAY for 24 h to activate the HIF transcriptional program prior to sample collection (**Fig \@ref(fig:m2)A**). Similar to hypoxia treatment, BAY decreased cell growth rate (**Figs \@ref(fig:m2)B-C**) and activated the HIF-1α transcriptional program (**Figs \@ref(fig:m1)D-H**). Compared to hypoxia, BAY treatment resulted in a similar activation of HIF-1α target gene transcription and protein expression. Unlike hypoxia, HIF-1α stabilization in normoxia markedly increased glucose uptake and lactate efflux (**Fig \@ref(fig:m1)I**), as expected based on the pattern of glycolytic gene expression that we observed. Interestingly, although hypoxia and BAY treatments resulted in similar increases in HIF-1α, GLUT1, and LDHA, the glycolytic response was markedly different.

## Extracellular flux measurements highlight subtle differences between hypoxia and BAY

In addition to glucose and lactate, we also determined the extracellular fluxes of pyruvate and amino acids (**Figs \@ref(fig:m1)J, \@ref(fig:m2)J**). Overall, changes were modest, with hypoxia generally decreasing the fluxes of all measured metabolites. Notably, a substantial decrease in glutamine consumption was observed in LFs cultured in 0.5% oxygen. This observation is in contrast to previous studies in cancer cell metabolism demonstrating increased glutamine uptake as a key feature of the metabolic response to hypoxia [@metallo2011; @wise2011; @gameiro2013]. Similar patterns were observed in 0.2% oxygen (**Figure S2**). In PASMCs, glutamine uptake did increase, as did the uptake of branched-chain amino acids and arginine (**Figure S3**), highlighting differential responses of these mesenchymal cells to hypoxia. Comparatively modest effects of BAY on amino acid fluxes were observed as compared to 0.5% oxygen culture conditions (**Figure \@ref(fig:m1)J**). BAY treatment elicited different responses to hypoxia with relative preservation of glutamine and asparagine uptake and alanine and glutamate efflux compared to hypoxia.

## Stable isotope labeling

>>>>

## Hypoxia uncouples HIF-dependent glycolytic gene expression from glycolytic metabolic flux

The goal of this study was to characterize hypoxia-induced metabolic changes in proliferating primary LFs and PASMCs. To accomplish this goal, we adopted a metabolic flux analysis (MFA) technique that enabled us to link intracellular metabolic fluxes to cell proliferation rates. MFA fits cell proliferation rate, extracellular flux measurements, and ^13^C intracellular isotope labeling patterns to a computational model of cell metabolism [@antoniewicz2018]. This analysis reconstructs comprehensive flux maps that depict the flow of carbon from extracellular substrates, through intracellular metabolic pathways, and into cell biomass and metabolic by-products [@young2014]. These models assume that cells are at a metabolic pseudo-steady state over the experimental time course [@buescher2015]. Exponential growth phase is thought to reflect metabolic pseudo-steady state as cells in the culture steadily divide at their maximal condition-specific rate, provided nutrient supply does not become limiting [@buescher2015; @ahn2011]. Thus, we first set out to define experimental conditions to capture exponential growth phase in normoxic and hypoxic cultures.

@@

Given that hypoxia exposure did not increase glycolysis in LFs, we next wanted to determine how these cells responded to HIF-1α stabilization in normoxia. To accomplish this, LFs were treated with the PHD inhibitor molidustat (BAY, 10 μM) using a similar time course as our hypoxia experiments. Cells were treated with BAY for 24 h to activate the HIF transcriptional program prior to sample collection (**Fig \@ref(fig:m2)A**). As with hypoxia, BAY decreased cell growth rate (**Figs \@ref(fig:m2)B-C**) and activated the HIF-1α transcriptional program (**Figs \@ref(fig:m1)D-H**). Compared to hypoxia, BAY treatment resulted in a similar activation of HIF-1α target gene transcription and protein expression. Unlike hypoxia, however, HIF-1α stabilization in normoxia markedly increased glucose uptake and lactate efflux (**Fig \@ref(fig:m1)I**), as expected based on the pattern of glycolytic gene expression that we observed. Interestingly, although hypoxia and BAY treatments resulted in similar increases in HIF-1α, GLUT1, and LDHA, the glycolytic response was markedly different.

## Extracellular fluxes are treatment and cell-type dependent

In addition to glucose and lactate, we also determined the extracellular fluxes of pyruvate and amino acids (**Figs \@ref(fig:m1)J, \@ref(fig:m2)J**). To our knowledge, this is the first comprehensive extracellular flux profiling of key metabolic substrates in these primary cells. In LFs, overall, changes in these fluxes were modest, with hypoxia generally decreasing the fluxes of all measured metabolites. These findings were similar with 0.2% oxygen exposure (**Fig S2J**).

Notably, we observed a significant decrease in glutamine consumption in hypoxic LFs. This finding contrasts with previous studies of cancer cell metabolism demonstrating increased glutamine uptake as a key feature of the metabolic response to hypoxia [@metallo2011; @wise2011; @gameiro2013]. In these systems, glutamine-derived 2-oxoglutarate was reductively carboxylated by isocitrate dehydrogenase enzymes to generate citrate for lipogenesis. In addition, glutamine has been shown to support TCA cycling in hypoxia in a Burkitt lymphoma model [@le2012]. Unlike LFs, PASMCs did exhibit a trend toward increased glutamine uptake (**Figure S3J**), suggesting a greater reliance on these metabolic pathways in their adaptive response to hypoxia.

In LFs, among all of the measured amino acid fluxes, proline consumption uniquely increased (**Fig \@ref(fig:m1)J**). Hypoxia increases collagen expression in these cells [@liu2013a] and proline constitutes ~ 10% of the total amino acid content of collagens. Together, these data suggest an important contribution of extracellular proline to collagen production in hypoxic LFs as has been observed in other fibroblast cell lineages [@szoka2017].

In PASMCs, we observed increased consumption of the branched-chain amino acids (BCAAs) leucine and valine as well as arginine (**Figure S3J**), which was not observed in LFs. BCAAs are transaminated by branch chain amino transferase enzymes to branched chain α-keto acids (BCKAs). BCKAs are further metabolized to yield acyl-CoA derivatives for lipogenesis or oxidation [@mann2021; @crown2015]. Previous studies have shown that hypoxia up-regulates arginase expression in hypoxic PASMCs [@chen2009; @xue2017] to support polyamine and proline synthesis required for cell proliferation [@li2001]. Interestingly, activation of these metabolic pathways in hypoxia was not observed in LFs and suggests distinct metabolic vulnerabilities of these different cell types.

Compared to hypoxia treatment, BAY demonstrated more modest effects on amino acid fluxes generally (**Figure \@ref(fig:m2)J**). In particular, glutamate efflux was not affected by BAY treatment, while it was reduced by hypoxia. Alanine efflux was increased by BAY treatment, but decreased by hypoxia. In addition to the glucose and lactate fluxes noted above, these findings further highlight fundamental differences in the metabolic consequences of HIF activation in normoxia and hypoxia.

## Isotope tracing demonstrates less incorporation of glucose into downstream pathways

To investigate intracellular metabolic reprogramming in hypoxic cells, we performed ^13^C stable isotope tracing with [U-^13^C~6~]-glucose, [1,2-^13^C~2~]-glucose, and [U-^13^C~5~]-glutamine. Isotopic enrichment of downstream metabolites in glycolysis and the TCA cycle were determined by LC-MS.
